---
title: "Transmission Type and Fuel Efficiency"
author: "John James"
date: "April 15, 2016"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: journal
  word_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gvlma)
library(knitr)
library(kfigr)
library(printr)
library(car)
read_chunk("mt.R")
knitr::opts_chunk$set(fig.path = "../figures/", fig.height = 4, fig.width = 10, echo=TRUE, warning=FALSE)
```

```{r environment, echo = FALSE, message = F, eval = T}
<<environment>>
```

```{r data, echo = FALSE, message = F, eval = T}
<<data>>
```

```{r corrOutcome, echo = FALSE, message = F, eval = T}
<<corrOutcome>>
```

```{r corrPredictors, echo = FALSE, message = F, eval = T}
<<corrPredictors>>
```

```{r modelSets, echo = FALSE, message = F, eval = T}
<<modelSets>>
```

```{r getPvalue, echo = FALSE, message = F, eval = T}
<<getPvalue>>
```

```{r plots, echo = FALSE, message = F, eval = T}
<<plots>>
```

```{r outliers, echo = FALSE, message = F, eval = T}
<<outliers>>
```

```{r collinearity, echo = FALSE, message = F, eval = T}
<<collinearity>>
```


```{r bootstrap, echo = FALSE, message = F, eval = T}
<<bootstrap>>
```

#Synopsis
The purpose of this analysis was to explore the 1974 Motor Trend Magazine car fuel consumption data set to determine whether automatic or manual transmissions were better for miles per gallon (mpg) and to quantify the difference, if any. A thorough and methodical multiple regression model selection process was undertaken to identify covariates that, together with transmission type, provided the most robust predictions of mpg. Regression techniques were employed to isolate and quantify the effect of transmission type and model diagnostics validated linear regression assumptions. This analysis concludes that transmission type, one of several predictive factors, had no statistically significant effect on mpg during the Motor Trend study and as such, **this analysis to determine the effect of transmission type on mpg is inconclusive**.

#Data
There were `r mtStats[5,1]` observations of 11 variables in the Motor Trend Cars dataset, 6 continuous, 5 categorical. `r figr("carsFormat", TRUE, type="Table")` outlines the structure of the dataset and `r figr("mtStats", TRUE, type="Table")` presents the summary statistics for the continuous variables.  MPG ranged from `r mtStats[3,1]` to `r mtStats[4,1]`, with a mean of `r round(mtStats[1,1],2)` mpg. Weight, measured in 1000 lb, ranged from `r round(1000 * mtStats[3,5],2)` to `r round(1000 * mtStats[4,5],2)` lb, with a mean of `r round(1000 * mtStats[1,5],2)` lb.  Mean horsepower and displacement was `r round(mtStats[1,3],2)` and `r round(mtStats[1,2],2)` cubic inches, respectively.  

#Method
**Relationship Evaluation**
The scatterplot in `r figr("scatterPlot1", TRUE, type="Figure")` depicts the relationships among mpg, transmission type (am), and the variables with which mpg was most highly correlated. Several observations were made.  First, the factors most correlated with mpg were weight, cylinders, displacement, and horsepower.  Second, displacement, horsepower and weight had a slightly curved relationship to mpg, suggesting a log transformation might improve linearity. Lastly, the degree of correlation among displacement, horsepower, cylinders, and weight, signaled a high potential for multicollinearity.

**Univariate Analysis**
A simple linear regression was conducted to predict mpg based on transmission type. The linear equation produced an adjusted R^2^ of `r round(summary(lm1)$adj.r.squared,3)` (F(`r summary(lm1)$fstatistic[2]`, `r summary(lm1)$fstatistic[3]`)=`r round(summary(lm1)$fstatistic[1],3)`, p < $`r round(lm1p,4)`$).  Though transmission type was found to be significant ($\beta_0 = `r summary(lm1)$coef[1,4]`$ and $\beta_1 = `r summary(lm1)$coef[2,4]`)$, the low R^2^ suggested other variables accounted for much of the variance in mpg. 

**Model Selection**
The regression equation for the full model produced an adjusted R^2^ of `r round(summary(lm10)$adj.r.squared,3)` (F(`r summary(lm10)$fstatistic[2]`, `r summary(lm10)$fstatistic[3]`)=`r round(summary(lm10)$fstatistic[1],3)`, p < $`r round(lm10p,4)`$).  However, the exceedingly high multicollinearity indicated by variance inflation factors which ranged from `r round(viflm10[1,1],3)` for `r rownames(viflm10)[1]` to `r round(viflm10[5,1],3)` for `r rownames(viflm10)[5]`, disqualified this model from selection.

The four predictors most highly correlated with fuel efficiency - weight, cylinders, displacement, and horsepower (`r figr("scatterPlot1", TRUE, type="Figure")`) - were added to the univariate model, one-by-one. The anova test in `r figr("anova1", TRUE, type="Table")` reveals significant effects for transmission, weight, cylinders, and horsepower.  Due to a lack of significance and high correlation with the other factors, displacement and cylinders were dropped from the model. Transmission type, modeled with the logs of weight and horsepower to improve linearity, produced the most robust predictions. The remaining independent variables were added to the model, one-by-one, in order of correlation with mpg. None were statistically significant contributors to mpg variation (`r figr("anova2", TRUE, type="Table")`). `r figr("target", TRUE, type="Table")` summarizes the target regression model coefficients, R^2^ and fit.

#Results
**Regression Model**
The selected regression model was: $Y_i = \beta_0 + \beta_1am_i + \beta_2log(wt_i) + \beta_3log(hp_i) + \epsilon_i$
where:

* *$\beta_0$* = mean mpg for automatic car with weight and horsepower = 0
* *$\beta_1$* = increase in mpg for manual transmission 
* *$\beta_2$* = increase (reduction) in mpg for each unit log of 1000 lb of weight 
* *$\beta_3$* = increase (reduction) in mpg for each unit log of hp

A significant regression equation was found (F(`r summary(lm72)$fstatistic[2]`, `r summary(lm72)$fstatistic[3]`)=`r round(summary(lm72)$fstatistic[1],3)`, p < $`r lm72p`$), with an adjusted R^2^ of `r round(summary(lm72)$adj.r.squared,3)`. Predicted mpg was equal to `r round(summary(lm72)$coef[1,1],3)`  + `r round(summary(lm72)$coef[2,1],3)` * (Transmission)  `r round(summary(lm72)$coef[3,1],3)` * (wt)  `r round(summary(lm72)$coef[4,1],3)` * (hp), where Transmission was coded as 0 = Auto, 1 = Manual, wt was the weight of the vehicle measured in 1000lb, and hp was gross horsepower. Miles per gallon increased by `r round(summary(lm72)$coef[2,1],3)` mpg for cars with manual transmissions, decreased by `r -round(summary(lm72)$coef[3,1],3)` mpg for each log 1000 lb of weight, and decreased by `r -round(summary(lm72)$coef[4,1],3)` mpg for each log unit of horsepower. Though log weight and log horsepower were significant predictors of mpg with p-values equal to $`r round(summary(lm72)$coef[3,4],4)`$ and $`r round(summary(lm72)$coef[4,4],4)`$ respectively, transmission type was not.  The p-value for transmission type was `r round(summary(lm72)$coef[2,4],3)`, significantly above our significance factor of α = 0.05. 
**Since transmission type was not a statistically significant predictor of mpg, this analysis to determine the effect of transmission type on mpg is inconclusive.**

**Regression Model Assumptions**
The linear regression basic assumptions hold for this model.

* *Existence:* The relationship between dependent and independent variables exists.  The model predictors produce an adjusted R^2^ of `r round(summary(lm72)$adj.r.squared,3)`, thereby explaining `r 100 * round(summary(lm72)$adj.r.squared,3)` percent of the variability in mpg.
* *Linearity:* The Residual vs Fitted plot shown in `r figr("plotlm72", TRUE, type="Figure")` evinces no distinct pattern among the residuals over the spectrum of values studied.  As such, the assumption of linearity holds for this model.
* *Normality:* The Normal Q-Q plot in `r figr("plotlm72", TRUE, type="Figure")` indicates that the distribution of standardized residuals approximates that of a sample of independent normals.  Therefore, the assumption of normality is valid for this model.
* *Constant Variance:*  The Scale-Location plot in `r figr("plotlm72", TRUE, type="Figure")` shows a collection of  square roots of standardized residuals dispersed evenly about a horizontal line.  The assumption of constant variance over the given values of the explanatory variables, is supported.

**Regression Model Validity**

* *Outliers:* The `r cd[1,2]`, `r cd[2,2]`, `r cd[3,2]`, and `r cd[4,2]` had the highest influence on the regression model with cook's distances of `r round(cd[1,1],3)`, `r round(cd[2,1],3)`, `r round(cd[3,1],3)`, and `r round(cd[4,1],3)`, respectively.  All points were lower than the cut-off value of $D_i > 1$ as indicated in the Residuals vs. Leverage plot in `r figr("plotlm72", TRUE, type="Figure")`. That said, omitting those points from the regression analysis resulted in a `r pctchg ` increase in adjusted R^2^, from `r round(summary(lm72)$adj.r.squared,3)` to `r round(summary(lm72_2)$adj.r.squared,3)`. This represents `r pctrem` percent of remaining variance - a material increase in model fit. **Notwithstanding, transmission type retained low significance as a predictor of mpg, with a p-value of `r round(summary(lm72_2)$coef[2,4],3)`.**  

* *Multicollinearity:* The variance inflation factors for transmission type, log weight, and log horsepower, were `r round(viflm72[1],3)`, `r round(viflm72[2],3)`, and `r round(viflm72[3],3)` respectively. This is sufficiently low to conclude that multicollinearity is not a factor in this model.

* *Bootstrap* To test the validity of the model, 1000 bootstrap samples were simulated and the sampling distribution of the model coefficients (statistic) under the null hypothesis $H_0$, was ascertained. An $H_0$ in which the hypothesized model statistics were the same as the sampling distribution was not rejected since all hypothesized coefficients were within the 95% confidence interval for the statistic `r figr("bsum", TRUE, type="Table")`.

#Appendix

##Figures
```{r scatterPlot1, echo = FALSE, message = F, eval = T, anchor = "Figure", fig.height = 6}
print(sp1)
```
`r figr("scatterPlot1", TRUE, type="Figure")`: Scatterplot Matrix 


```{r plotlm72, echo = FALSE, message = F, eval = T, anchor = "Figure", fig.height = 6}
layout(matrix(c(1:4),2,2))
plot(lm72, which = c(1,2,3,5))
```
`r figr("plotlm72", TRUE, type="Figure")`: Model 3: Target Model Fit Analysis

##Tables
`r figr("carsFormat", TRUE, type="Table")`: Format of Motor Trend Cars Data Set
```{r carsFormat, echo = FALSE, message = F, eval = T, anchor = "Table"}
carsFormat
```

`r figr("mtStats", TRUE, type="Table")`: Summary Statistics for Motor Trend Cars Continuous Variables
```{r mtStats, echo = FALSE, message = F, eval = T, anchor = "Table"}
mtStats
```

`r figr("anova1", TRUE, type="Table")`: Model Series 1 Coefficient Significance Test
```{r anova1, echo = FALSE, message = F, eval = T, anchor = "Table"}
print(anova(lm1, lm2, lm3, lm4, lm5))
```

`r figr("anova2", TRUE, type="Table")`: Model Series 2 Coefficient Significance Test
```{r anova2, echo = FALSE, message = F, eval = T, anchor = "Table"}
print(anova(lm72, lm73, lm74, lm75, lm76, lm77))
```

`r figr("target", TRUE, type="Table")`: Summary of Target Regression Model
```{r target, echo = FALSE, message = F, eval = T, anchor = "Table"}
summary(lm72)
```

`r figr("bsum", TRUE, type="Table")`: Bootstrap Summary and Confidence Interval
```{r bsum, echo = FALSE, message = F, eval = T, anchor = "Table"}
bsum
```

